Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-10-30 ms/ms.Rmd
add Skrimfjella occurrence evaluation

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019]. 
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.

@@

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space [@brusSamplingDigitalSoil2019].
Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used Random Forests (RF) to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

@@

Extrapolation.

>>>>

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019]. 
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space [@brusSamplingDigitalSoil2019].
Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used RF to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

@@

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first --- for both sites --- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.

### 2024-10-30 ms/ms.tex
add Skrimfjella occurrence evaluation

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.

@@

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used Random Forests (RF) to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

@@

Extrapolation.

>>>>

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used RF to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

@@

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first --- for both sites --- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
